Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Matter Power Spectra

### 2025-06-30 paper/main.tex
fix typos, minor adjustments

\begin{abstract} 
Understanding the structure of our universe and the distribution of matter is an 
area of active research.  As cosmological surveys grow in complexity, the development 
of emulators to efficiently and effectively predict matter power spectra is essential.  
We are particularly motivated by the Mira-Titan Universe simulation
suite which, for a specified cosmological parameterization (termed a ``cosmology''), 
provides multiple response curves of various fidelities, including correlated 
functional realizations.  Our objective is two-fold.  First, we estimate 
the underlying true matter power spectra, with appropriate uncertainty 
quantification (UQ), from all of the provided curves.  To this end, we propose a 
novel Bayesian deep Gaussian process (DGP) hierarchical model which synthesizes 
all the simulation information to estimate the underlying matter power spectra
while providing effective UQ.  Our model extends previous work on Bayesian DGPs 
from scalar responses to correlated functional outputs.  Second, we leverage our predicted 
power spectra from various cosmologies in order to accurately predict the entire 
matter power spectra for an 
unobserved cosmology.  For this task, we use basis function representations 
of the functional spectra to train a separate Gaussian process emulator.  
Our method performs well in synthetic exercises and against the benchmark cosmological 
emulator (Cosmic Emu).
\end{abstract}

@@

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,derose2019aemulus,
nishimichi2019dark,angulo2021bacco,euclid2021euclid,moran2023mira}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectra behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
Our overarching goal is to leverage such simulation data to 
first estimate and then predict underlying matter power spectra for various 
cosmological parameterizations.

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the matter power spectrum 
is rather complex.  Although we presume there exists some true matter spectrum as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulation returns
a convoluted patchwork of potential spectra, which vary in fidelity and are only trusted
over particular wavenumber ranges.  Our first objective is to synthesize expert knowledge
and all available simulation information into an effective estimate of the true underlying
matter spectrum.

There are several challenges to estimating the underlying spectrum.  Our model
must: handle multiple correlated functional observations, offer enough 
flexibility to accommodate the nonstationarity present in matter power spectra 
(which vary in smoothness across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model which treats functional simulation 
outputs as realizations of a Gaussian process (GP) 
centered on the true underlying spectrum, which itself is a realization of a deep Gaussian process 
\citep[DGP;][]{damianou2013deep}.  The GP accommodates correlated observations
through its covariance structure.  The depth of the DGP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
To demonstrate the profiency of our Bayesian hierarchical DGP, 
we benchmark its performance in two unique settings: with synthetic data 
mimicing the Mira-Titan data and with real-world data from 
the Code for Anisotropies in the Microwave Background \citep[CAMB;][]{lewis2011CAMB}.
The CAMB model exhibits similar behavior to Mira-Titan but crucially
provides a ``true" power spectrum for each cosmology.

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict matter power spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
(one batch of simulations can take multiple weeks to run on a supercomputer),
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration which may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} which will provide quick and effective
predictions of matter power spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 111 simulations, and is termed 
the ``Cosmic Emu.''\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a GP surrogate 
on principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

@@

For each cosmology, Mira-Titan contains a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
low resolution simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$). 
These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), the spatially correlated errors of the low-resolution and high-resolution runs 
becomes apparent.

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c\approx 3.73$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. 
We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\begin{equation}\label{eq:lambda}
\Lambda_p^{(ii)} = \begin{cases}
    10^8 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\end{equation}
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^8$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectrum, the weighted average fluctuates too drastically.  
This is especially apparent after subtracting a LOESS-smoothed weighted average (right panel).
Although the weighted average incorporates all available observatons and 
expert knowledge regarding the precisions, it fails to account for the spatial 
dependence inherent in the smooth but stochastic spectra, which is pivotal to 
effectively inferring the true spectrum.

@@

Let $S$ represent the true (but unknown) underlying matter power spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2023mira}, we find it
essential to account for the smoothly varying spatial dependence
(as visualized in Figure \ref{fig:plot_data}).  We thus specify a dense 
covariance matrix $\Sigma_\varepsilon$ that simulataneously
allows us to model the correlated errors and build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

@@

Our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, but this requires inferring latent $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~(\ref{eq:dgphm}) into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{monoGP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
A detailed derivation, including generalizations for 
$\boldsymbol{\mu}_S\neq\mathbf{0}$, is provided in \ref{sec:apdx_int_lik}. 
The incorporation of $\Sigma_\epsilon$ in the
Gaussian likelihood of this outer layer is an essential upgrade to previous DGP inferential schemes.
This covariance incorporates knowledge of the smoothly varying dynamics from which $\bar{Y}$ was
generated.

@@

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2023mira}, we consider two test functions 
defined as
\[
\begin{aligned}
  f_1(x) &= m_1 \exp\left(-\frac{u_1x}{2}\right) * 
  \cos\left(x\sqrt{25-\left(\frac{u_1}{2}\right)^2}\right) - \frac{m_1x}{5}\\
  f_2(x) &= \exp\left(-m_2(x-3)^2\right) + 
  \exp\left(-u_2(x-1)^2\right)-0.05\sin\left(8(x-1.9)\right).
\end{aligned}
\]
For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are randomly selected from
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$. The outputs of $f_1$ and $f_2$ (conditional on $m_1/u_1$ and 
$m_2,u_2$), determine the true function values for each simulation.

@@

\[
\begin{aligned}
\Sigma_A &= 0.0225 * K(||x_i - x_j||^2, 0.01), \\
\Sigma_B &= s^\top Ms \quad\textrm{where}\quad M = 0.1 * K(||x_i - x_j||^2, 0.05)
\quad\textrm{and}\quad s = 1.5^{-x/2}.
\end{aligned}
\]
Observations are kept noise-free, but we do add a jitter of $10^{-8}$ to the diagonal of $\Sigma_A$ 
and $\Sigma_B$ for numerical stability.  Within $\Sigma_B$, $s$ introduces a degree of
nonstationarity, more effectively mimicing the behavior of the 
CAMB and Mira-Titan datasets.  Examples of simulation output are showin in 
Figure \ref{fig:run_sim_f1A_f2B}.

@@

For each simulated function and variance specification, we conduct 20 Monte Carlo repetitions
with re-randomized $m_1, u_1, m_2, u_2$.  We consider $r = 5$ and $r = 15$.
We compare to an out-of-the-box DGP from the 
\texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP from
the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, and
a deep process convolution (DPC) approach, which is utilized within Cosmic 
Emu \citep{moran2023mira}.  These competing methods view the training data as 
noisy observations with no spatial dependence (i.e., they view the gray dots of
Figure \ref{fig:run_sim_f1A_f2B} without the lines). In
contrast, our method leverages the fact that the observations represent $r$-many
smooth functional realizations in order to estimate parameters
within $\Sigma_\epsilon$ that account for the spatial dependence.
Our method achieves this by first pre-scaling the functional realizations by the
precision values (to account for nonstationarity), and then using maximum likelihood
estimation to obtain estimates of the marginal variance and lengthscale of the 
Mat\'ern spatial correlation function. We use these estimates and the precision 
information to construct $\Sigma_\varepsilon$.

@@

We then fit our DGP.FCO model on each of 111 training cosmologies, reserving 6 cosmologies
for later hold-out testing.  Figure \ref{fig:plot_fit} 
visualizes the DGP.FCO fit for the first training cosmology.  We center the plot
by subtracting the posterior mean in order to more easily visualize the credible
interval bounds.  The UQ from the DGP.FCO model accounts for varying precisions 
across differing $k$ values, as well as the spatial correlation within the 
functional realizations.  To provide insight into the nonstationarity of the power
matter spectrum, the right panel of Figure \ref{fig:plot_fit} shows burned-in elliptical 
slice samples of $W$.
These warpings depart from the identiy mapping around $x\approx-1$, 
stretching inputs where dynamics are shifting more rapidly for larger $k$ values.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_dot.jpeg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means and training with the ``true" infinite-resolution spectra. The identity line 
             provides equal MSE values amongst both methods.}   
    \label{fig:mse_camb}
\end{figure}

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=\textwidth]{pred_1to6.jpeg}
    \caption{Results of all six predictions for both methods on the held-out 
    cosmologies (DGP.FCO posterior mean subtracted). CosmicEMU is dotted blue and DGP.FCO is dashed orange.
    Closer to the dotted zero line indicates a better fit with the DGP.FCO posterior mean.}
    \label{fig:plot_pred_1to6}
\end{figure}

@@

This work was in part supported by the U.S. Department
of Energy, Office of Science, Office of Advanced Scientific Computing Research, 
Scientific Discovery through Advanced Computing (SciDAC) program through 
Grant 420453 ``Enabling Cosmic Discoveries in the Exascale Era."

>>>>

\begin{abstract} 
Understanding the structure of our universe and the distribution of matter is an 
area of active research.  As cosmological surveys grow in complexity, the development 
of emulators to efficiently and effectively predict matter power spectra is essential.  
We are particularly motivated by the Mira-Titan Universe simulation
suite that, for a specified cosmological parameterization (termed a ``cosmology''), 
provides multiple response curves of various fidelities, including correlated 
functional realizations.  Our objective is two-fold.  First, we estimate 
the underlying true matter power spectra, with appropriate uncertainty 
quantification (UQ), from all of the provided curves.  To this end, we propose a 
novel Bayesian deep Gaussian process (DGP) hierarchical model which synthesizes 
all the simulation information to estimate the underlying matter power spectra
while providing effective UQ.  Our model extends previous work on Bayesian DGPs 
from scalar responses to correlated functional outputs.  Second, we leverage our predicted 
power spectra from various cosmologies in order to accurately predict the entire 
matter power spectra for an 
unobserved cosmology.  For this task, we use basis function representations 
of the functional spectra to train a separate Gaussian process emulator.  
Our method performs well in synthetic exercises and against the benchmark cosmological 
emulator (Cosmic Emu).
\end{abstract}

@@

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments that simulate the expansion of the universe are growing
in prevalence and complexity \citep[e.g.,][]{lawrence2010coyote,derose2019aemulus,
nishimichi2019dark,angulo2021bacco,euclid2021euclid,moran2023mira}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectra behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
Our overarching goal is to leverage such simulation data to 
first estimate and then predict underlying matter power spectra for various 
cosmological parameterizations.

The Mira-Titan simulation suite \citep{moran2023mira}
defines ``cosmologies'' according to eight cosmological parameters (more on this
in Section \ref{sec:data}).  For a particular cosmology, the simulation of the matter power spectrum 
is rather complex.  Although we presume there exists some true matter spectrum as a function
of wavenumber, we are not able to directly observe it.  Rather, the simulations for a 
particular cosmology return a collection of multiple potential spectra, 
which vary in fidelity and are only trusted over particular wavenumber ranges.  
Our first objective is to synthesize expert knowledge and all available 
simulation information into an effective estimate of the true underlying
matter spectrum.

There are several challenges to estimating the underlying spectrum.  Our model
must: handle multiple correlated functional observations, offer enough 
flexibility to accommodate the nonstationarity present in matter power spectra 
(which vary in smoothness across wavenumbers), provide effective uncertainty
quantification (UQ), and allow for the incorporation of expert knowledge 
regarding the wavenumbers over which various outputs are trusted.  To this end, 
we propose a Bayesian hierarchical model that treats functional simulation 
outputs as realizations of a Gaussian process (GP) 
centered on the true underlying spectrum, which itself is a realization of a deep Gaussian process 
\citep[DGP;][]{damianou2013deep}.  The GP accommodates correlated observations
through its covariance structure.  The depth of the DGP offers nonstationary
flexibility.  The Bayesian framework facilitates UQ and the incorporation of prior
knowledge.  While Bayesian DGPs have been previously deployed for computer experiments 
with scalar responses \citep[e.g.,][]{sauer2023active,sauer2023vecchia,ming2023deep}, 
they have yet to be extended to correlated functional outputs. 
To demonstrate the proficiency of our Bayesian hierarchical DGP, 
we benchmark its performance in two unique settings: with synthetic data 
mimicking the Mira-Titan data and with predictions of the linear power spectrum from 
the Code for Anisotropies in the Microwave Background \citep[CAMB;][]{lewis2011CAMB}.
The CAMB model exhibits similar behavior to Mira-Titan but crucially
provides a ``true" power spectrum for each cosmology.

Our second objective is to leverage simulation data from a limited set of
cosmologies in order to predict matter power spectra for unobserved cosmologies.
Due to the computationally expensive nature of the Mira-Titan simulation suite
(one batch of simulations can take multiple weeks to run on a supercomputer),
it is infeasible to evaluate the simulation for every possible eight-dimensional
cosmological configuration that may be of interest.  Instead, we desire
a statistical ``emulator'' or ``surrogate model'' 
\citep{santner2003design,gramacy2020surrogates} that will provide quick and effective
predictions of matter power spectra for any cosmological parameterization.
\citet{moran2023mira} provide a state-of-the-art emulator for the Mira-Titan
simulation suite, which was trained on 111 simulations, and is termed 
the ``Cosmic Emu.''\footnote{\url{https://github.com/lanl/CosmicEmu}}
We leverage the same training data, in conjunction with our Bayesian hierarchical
DGP model and basis function representations, to train a GP surrogate 
on principal component weights in order to predict spectra for unobserved
cosmologies \citep{higdon2008computer, higdon2010estcosmo}. 
Our method compares favorably to Cosmic Emu on held-out cosmologies.

@@

For each cosmology, Mira-Titan contains a batch of 18 simulated power spectra: an inexpensive 
power spectrum estimated from perturbation theory ($y_p$), 16 power spectra estimated from 
low resolution simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one power spectrum from 
a high resolution simulation ($y_h$). 
These power spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), the spatially correlated fluctuations of the low-resolution and 
high-resolution runs becomes apparent.

Notice, in the center/right panels of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectrum increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c\approx 3.73$ for 
the increase in precision from the low- to high-resolution output is also 
estimated with this regression model. 
We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the $n\times n$ diagonal matrices with elements
\begin{equation}\label{eq:lambda}
\Lambda_p^{(ii)} = \begin{cases}
    10^8 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\end{equation}
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^8$ to impose the deterministic nature of the perturbation theory output.
In $\Lambda_\ell$, the multiplication by 16 accounts for the sample size of the
low resolution runs.
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectrum, the weighted average fluctuates too drastically.  
This is especially apparent after subtracting a LOESS-smoothed weighted average (right panel).
Although the weighted average incorporates all available observations and 
expert knowledge regarding the precisions, it fails to account for the spatial 
dependence inherent in the smooth but stochastic spectra, which is pivotal to 
effectively inferring the smooth spectrum.

@@

Let $S$ represent the true (but unknown) underlying matter power spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2023mira}, we find it
essential to account for the smoothly varying spatial dependence
(as visualized in Figure \ref{fig:plot_data}).  We thus specify a dense 
covariance matrix $\Sigma_\varepsilon$ that simultaneously
allows us to model the correlated errors and build domain-matter expertise
into the prior. To this end, we embrace variations of the Mat\'ern kernel 
\citep{stein1999interpolation}, tailored to each cosmology (more on this in later 
subsections).

@@

Our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, but this requires inferring latent $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~(\ref{eq:dgphm}) into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{monoGP}\left(\mathbf{0}, \Sigma_W(X)\right).
\end{align}
A detailed derivation, including generalizations for 
$\boldsymbol{\mu}_S\neq\mathbf{0}$, is provided in \ref{sec:apdx_int_lik}. 
The incorporation of $\Sigma_\epsilon$ in the
Gaussian likelihood of this outer layer is an essential upgrade to previous DGP inferential schemes.
This covariance incorporates knowledge of the smoothly varying dynamics from which $\bar{Y}$ was
generated.

@@

We compare DGP.FCO to competing models with a simulation study where the underlying 
true curve is known. Following \cite{moran2023mira}, we consider two test functions 
defined as
\[
\begin{aligned}
  f_1(x) &= m_1 \exp\left(-\frac{u_1x}{2}\right) * 
  \cos\left(x\sqrt{25-\left(\frac{u_1}{2}\right)^2}\right) - \frac{m_1x}{5}\\
  f_2(x) &= \exp\left(-m_2(x-3)^2\right) + 
  \exp\left(-u_2(x-1)^2\right)-0.05\sin\left(8(x-1.9)\right).
\end{aligned}
\]
For additional variation, the values of $m_1, m_2, u_1,$ and $u_2$ are randomly selected from
$m_1 \sim \text{Uniform}(0.5,1.5)$, $u_1 \sim \text{Uniform}(1.5,2.5)$, and $m_2,u_2 
\sim \text{Uniform}(0.6,1.4)$. The outputs of $f_1$ and $f_2$ (conditional on $m_1,u_1$ and 
$m_2,u_2$), determine the true function values for each simulation.

@@

\[
\begin{aligned}
\Sigma_A &= 0.0225 * K(||x_i - x_j||^2, 0.01), \\
\Sigma_B &= s^\top Ms \quad\textrm{where}\quad M = 0.1 * K(||x_i - x_j||^2, 0.05)
\quad\textrm{and}\quad s = 1.5^{-x/2}.
\end{aligned}
\]
Observations are kept noise-free, but we do add a jitter of $10^{-8}$ to the diagonal of $\Sigma_A$ 
and $\Sigma_B$ for numerical stability.  Within $\Sigma_B$, $s$ introduces a degree of
nonstationarity, more effectively mimicking the behavior of the 
CAMB and Mira-Titan datasets.  Examples of simulation output are shown in 
Figure \ref{fig:run_sim_f1A_f2B}.

@@

For each simulated function and variance specification, we conduct 20 Monte Carlo repetitions
with re-randomized $m_1, u_1, m_2, u_2$.  We consider $r = 5$ and $r = 15$.
We compare to an out-of-the-box DGP from the 
\texttt{deepgp} package \citep{sauer2023active}, a heteroskedastic GP from
the \texttt{hetGP} package \citep{binois2018practical, binois2021hetgp}, and
a deep process convolution (DPC) approach, which is utilized within the Cosmic 
Emu \citep{moran2023mira}.  These competing methods view the training data as 
noisy observations with no spatial dependence (i.e., they view the gray dots of
Figure \ref{fig:run_sim_f1A_f2B} without the lines). In
contrast, our method leverages the fact that the observations represent $r$-many
smooth functional realizations in order to estimate parameters
within $\Sigma_\epsilon$ that account for the spatial dependence.
Our method achieves this by first pre-scaling the functional realizations by the
precision values (to account for nonstationarity), and then using maximum likelihood
estimation to obtain estimates of the marginal variance and lengthscale of the 
Mat\'ern spatial correlation function. We use these estimates and the precision 
information to construct $\Sigma_\varepsilon$.

@@

We then fit our DGP.FCO model on each of 111 training cosmologies, reserving 6 cosmologies
for later hold-out testing.  Figure \ref{fig:plot_fit} 
visualizes the DGP.FCO fit for the first training cosmology.  We center the plot
by subtracting the posterior mean in order to more easily visualize the credible
interval bounds.  The UQ from the DGP.FCO model accounts for varying precisions 
across differing $k$ values, as well as the spatial correlation within the 
functional realizations.  To provide insight into the nonstationarity of the power
matter spectrum, the right panel of Figure \ref{fig:plot_fit} shows burned-in elliptical 
slice samples of $W$.
These warpings depart from the identity mapping around $x\approx-1$, 
stretching inputs where dynamics are shifting more rapidly for larger $k$ values.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mse_dot.jpeg}
    \caption{Comparison of MSE for CAMB predictions, when training with DGP.FCO posterior 
             means and training with the ``true" infinite-resolution spectra. The identity line 
             provides equal MSE values among both methods.}   
    \label{fig:mse_camb}
\end{figure}

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=\textwidth]{pred_1to6.jpeg}
    \caption{Results of all six predictions for both methods on the held-out 
    cosmologies (DGP.FCO posterior mean subtracted). Cosmic Emu is dotted blue and DGP.FCO is dashed orange.
    Closer to the dotted zero line indicates a better fit with the DGP.FCO posterior mean.}
    \label{fig:plot_pred_1to6}
\end{figure}

@@

This work was in part supported by the U.S. Department
of Energy, Office of Science, Office of Advanced Scientific Computing Research, 
Scientific Discovery through Advanced Computing (SciDAC) program through 
Grant 420453 ``Enabling Cosmic Discoveries in the Exascale Era." Work at Argonne 
National Laboratory was supported under the U.S. Department of Energy 
contract DE-AC02-06CH11357.
